The Economics of Shotgun Marriage

### 2020-10-02 main/draft.tex
Current stage

\documentclass[12pt,letter]{article}
\usepackage[left=0.8in,right=0.8in,top=1in,bottom=1in]{geometry}
\usepackage{amsmath}
\usepackage{pdflscape}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{multicol}
\usepackage{multirow}
\usepackage{microtype}
\usepackage{euscript}
\usepackage{epsfig}
\usepackage{epstopdf}
\usepackage{mathrsfs}
\usepackage{tikz}
\usepackage{verbatim}
\usetikzlibrary{shapes,arrows}
\usetikzlibrary{positioning}
\tikzstyle{block} = [rectangle, draw, rounded corners]
\tikzstyle{line} = [draw, -latex']
\newcommand{\hypo}{\mathcal{H}}            
\bibliographystyle{ieeetr}

@@

\newcommand{\bs}[1]{\boldsymbol{#1}}  
\newcommand{\bsA}{\boldsymbol{A}}

%\setstretch{1}                         
\flushbottom                            
\righthyphenmin=2                      
\pagestyle{plain}                       
%\settimeformat{hhmmsstime}  
\widowpenalty=300                   
\clubpenalty=3000                     
\setlength{\parindent}{0em}           
\setlength{\topsep}{0pt}              
\usepackage[pdftex,unicode,colorlinks=true,urlcolor=blue]{hyperref}
\usepackage{bbm}
\usepackage{tabularx}
\renewcommand{\emptyset}{\varnothing}

@@

\newcommand{\laseq}{\stackrel{\lambda\text{-a.e.}}{=}}
\renewcommand{\d}{\underline}
\renewcommand{\u}{\overline}
\newcommand{\td}{\underline{\theta}}
\newcommand{\tu}{\overline{\theta}}
%\renewcommand{\theenumi}{\alph{enumi}}
%\renewcommand{\labelenumi}{(\theenumi)}
%\renewcommand{\theenumii}{\roman{enumii}}
%\renewcommand{\labelenumii}{\theenumii.}
%\renewcommand{\theenumiii}{\arabic{enumiii}}
%\renewcommand{\labelenumiii}{\theenumiii.}
%\renewcommand{\epsilon}{\varepsilon} 
\newcommand{\hneq}{\stackrel{\hn}{=}}
\newcommand{\deq}{\stackrel{d}{=}}
\title{The Economics of Shotgun Marriage}
\author{Egor Kozlov\thanks{Economics Department, Northwestern University, egorkozlov2020@u.northwestern.edu. I thank my advisors Matthias Doepke, Martí Mestieri and David Berger for outstanding guidance and proofreading as part of my dissertation committee at Northwestern. I am grateful to Alessandra Voena for fantastic amount of help and support. I thank my colleagues Jane Olmstead-Rumsey, Kristina Manysheva, Bence Bardoczy, Stephanie Johnson and Gabriela Cugat, as well as all the participants of Macro Lunch at Northwestern for lots of helpful comments.}\\
{\small Northwestern University}}
\date{August, 2019}
\begin{document}
%\begin{center}\textbf{Economics 416-1} \\ \emph{By Egor Kozlov}\end{center}
\maketitle

\begin{abstract}
Many couples marry shortly after woman gets pregnant, which is often referred to as a shotgun marriage. I show that in American Community Survey data couples who experienced a shotgun marriage, i.e. had their first child before or at the year of their marriage, divorce more often in the future.
%This feature is robust in different subpopulations, being relatively more pronounced for wealthier people, although the share of such marriages is lower among them.
I argue that this difference in divorce rates can be explained by endogenous selection into marriage: women who experienced unplanned pregnancy agree to enter marriages of lower quality. 
This implies that in the data having kids before marriage is a plausible proxy for having lower-quality marriage.
Then I show that these patterns in selection can be rationalized by the structure of childcare costs, where kids require both money and time, and disproportional exposure of women to these costs.
I estimate a structural model endogenizing marriage selection and use it to argue that causal impact of unplanned fertility shock,
%differences in marriage quality,
as opposed to different composition of the groups, drives the major part of the observable difference in divorce chances.
Qualitatively, this suggests that the main reason for poor performance of shotgun marriages is large, unexpected and heterogeneous childcare costs. Quantitatively, the model suggests sizable effects of risks of unplanned pregnancy on observable divorce rates, especially for young women.
\end{abstract}

@@

This paper has two main goals. First, I document and quantify the persistent difference in divorce rates of couples who experienced shotgun marriages and how it is pronounced for different population groups. Second, based on these findings I estimate a structural model that is capable of identifying importance of different channels that drive this difference.
%In this paper I argue that this pattern is rationalized by partners agreeing to enter marriages of lower quality as a response to unplanned pregnancy shock.

%[Paragraph 2: general importance? Child development]
Understanding the sources of a variation in divorce rates is not just a household behavior question. For example, divorces have major impact on how people raise their children. The families who experienced shotgun marriages are often overlooked by the current literature focusing on child development. Growing up with a single mother or in a blended family (i.e. with step-parents) if conventionally considered\footnote{See Kearney, Levine 2017\nocite{kearney} for an extensive review of these issues} as a predictor of poor early and adult-life outcomes of the children, and those who have two married biological parents are typically not considered to be at risk. However, not all marriages are created equal, and treating all marriages equally may understate inequality: growing up in unstable marriage can be a predictor of ending up with a single mother or in a different household, as well as factor of risk by its own.

@@

%[Paragraph 7: model --- setup and why do we need a model]
These comparisons, however, contain lots of selection issues and unobservable heterogeneity, and to deal with it I set up a structural model that is capable of understanding of anatomy of the result. The model is based on a growing branch of literature in dynamic household bargaining, and endogenizes savings, marriage and divorce decisions, fertility decisions and fertility shocks, together with the structure of childcare costs. The last element is crucial in generating heterogeneous response in terms of income or education: as time of wealthier parents is more expensive, unexpected childcare costs create relatively more distress for them. The model is estimated to match the key patterns of the data coming from American Community Survey, as well as supplementary sources, including ATUS and CEX for disciplining childcare costs part, and the structure I use fits the data well.

%[Paragraph 8: model --- what generates the result in theory]
The main driving force in the model is an unplanned pregnancy shock. Each period single people meet potential partners with some probability, and if the partner is met, the couple decides whether to marry each other based on a simple bargaining procedure. The potential matches are heterogeneous in partner's characteristics and marriage quality, the latter is modeled as idiosyncratic joint utility gain of couple from being together, that can be interpreted as a ``love shock''.  Before the bargaining decision takes place, partners may or may not get an unplanned pregnancy shock, that proxies for unplanned children appearing in cohabitating unions as I do not model cohabitation explicitly. The shock interfers with bargaining: if the couple made a baby, the woman risks to stay a single mother if the couple does not agree to marry. The conjecture is that in matches that experienced the shock women agree to enter the marriages of lower quality. Importance and even presence of this channel, however, depend on parameters of individual preferences and childcare costs that I estimate structurally.

%[Paragraph 9: model --- what I get from it, mechanics and anatomy]
The key conclusion of the estimation part is that couples who experienced an unplanned pregnancy shock indeed agree to enter the marriages of the lower quality, and this is the major driver of the number that we observe in the data. The model is capable of delivering various decompositions that split observable cross-sectional difference in divorce rates onto effects of age, marriage length, income, wealth and marriage quality. As it turns out, the last factor is the most important contributor to the difference, and its contribution is [of the similar magnitude than the observable difference]. Decomposing the result further, the model predicts that the major part of the difference in divorce rates is driven by couples who would not have married if they did not have a baby (i.e. who complied with the shock), although there is some additional distress of having an unplanned baby on those who would have married anyway.

%[Paragraph 10: model --- counterfactuals and speculative things]
Additionally, the estimated model delivers predictions exploiting interaction between unplanned pregnancies and marriage quality. First, a major part of divorces that happen early in life are seen as driven by unexpected pregnancies, though divorces happening later are not. Second, the general impact of pressure of becoming a single mother on marriage decisions is substantial, and eliminating unexpected pregnancies happening before the marriage decision takes place lowers amount of people married by 25 by [...]. [Last, the model predicts decline in probabilities of unexpected pregnancies since 1980: the model matching fertility / marriage / divorce profiles from Census of 1980 predicts larger chances of unplanned pregnancies among all income and age groups.]

@@

%[Paragraph 12: what I learned about channels]
The main causal insight is that being a single mother is a major threat affecting how people make their marriage decisions, and large and unequally distributed costs of raising young children can be a driver in the creation of poor quality marriages. This channel can be of major importance for studying welfare policies subsidizing childcare, young families and single mothers, as well as for enforcing alimony payments and child support in general. This also emphasizes the link between contraception and abortion access and marriages, although this impact is limited as even the wealthiest couples do experience sizable share of shotgun marriages, though this may be cultural rather than economic issue.

@@

\section*{Literature}
\bibliography{bib}

>>>>

\documentclass[12pt,letter]{article}
\usepackage[left=0.8in,right=0.8in,top=1in,bottom=1in]{geometry}
\usepackage{amsmath}
\usepackage{pdflscape}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{multicol}
\usepackage{multirow}
\usepackage{microtype}
\usepackage{euscript}
\usepackage{epsfig}
\usepackage{epstopdf}
\usepackage{mathrsfs}
\usepackage{natbib}
\usepackage{tikz}
\usepackage{verbatim}
\usetikzlibrary{shapes,arrows}
\usetikzlibrary{positioning}

\tikzstyle{block} = [rectangle, draw, rounded corners]
\tikzstyle{line} = [draw, -latex']
\newcommand{\hypo}{\mathcal{H}}

@@

\newcommand{\bs}[1]{\boldsymbol{#1}}  
\newcommand{\bsA}{\boldsymbol{A}}
%\setstretch{1}                         
\flushbottom                            
\righthyphenmin=2                      
\pagestyle{plain}                       
%\settimeformat{hhmmsstime}  
\widowpenalty=300                   
\clubpenalty=3000                     
\setlength{\parindent}{0em}           
\setlength{\topsep}{0pt}              
\usepackage[pdftex,unicode,colorlinks=false,urlcolor=blue,citecolor=black,hidelinks]{hyperref}
\usepackage{bbm}
\usepackage{tabularx}
\renewcommand{\emptyset}{\varnothing}

@@

\newcommand{\laseq}{\stackrel{\lambda\text{-a.e.}}{=}}
\renewcommand{\d}{\underline}
\renewcommand{\u}{\overline}
\newcommand{\td}{\underline{\theta}}
\newcommand{\tu}{\overline{\theta}}
%\renewcommand{\theenumi}{\alph{enumi}}
%\renewcommand{\labelenumi}{(\theenumi)}
%\renewcommand{\theenumii}{\roman{enumii}}
%\renewcommand{\labelenumii}{\theenumii.}
%\renewcommand{\theenumiii}{\arabic{enumiii}}
%\renewcommand{\labelenumiii}{\theenumiii.}
%\renewcommand{\epsilon}{\varepsilon} 
\newcommand{\hneq}{\stackrel{\hn}{=}}
\newcommand{\deq}{\stackrel{d}{=}}
\title{The Economics of Shotgun Marriage}
\author{Egor Kozlov\thanks{Economics Department, Northwestern University, egorkozlov2020@u.northwestern.edu. I thank my advisors Matthias Doepke, Martí Mestieri and David Berger for outstanding guidance and proofreading as part of my dissertation committee at Northwestern. I am grateful to Alessandra Voena for fantastic amount of help and support. I thank my colleagues Jane Olmstead-Rumsey, Kristina Manysheva, Bence Bardoczy, Stephanie Johnson and Gabriela Cugat, as well as all the participants of Macro Lunch at Northwestern for lots of helpful comments. I owe special thanks to Fabio Blasutto for tons of help with both technical and substantial parts during his stay at Northwestern.}\\
{\small Northwestern University}}
\date{First Draft: May, 2019 \\[1em] This Draft: February, 2020}
\begin{document}
\bibliographystyle{apa}
%\begin{center}\textbf{Economics 416-1} \\ \emph{By Egor Kozlov}\end{center}
\maketitle

\begin{abstract}
Many couples marry shortly after woman gets pregnant, which is often referred to as a shotgun marriage. I show that in the US household data couples who experienced a shotgun marriage, i.e. had their first child before or at the year of their marriage, divorce more often in the future. I argue that this difference in divorce rates can be explained by endogenous selection into marriage: women who experienced unplanned pregnancy agree to enter marriages of lower quality. This implies that in the data having kids before marriage is a plausible proxy for having lower-quality marriage. Then I show that these patterns in selection can be rationalized by the structure of childcare costs, where kids require both money and time, and disproportional exposure of women to these costs. I estimate a structural model endogenizing marriage selection and use it to argue that causal impact of unplanned fertility shock, as opposed to different composition of the groups, drives the major part of the observable difference in divorce chances. Qualitatively, this suggests that the main reason for poor performance of shotgun marriages is large, unexpected and heterogeneous childcare costs. Quantitatively, the model suggests sizable effects of risks of unplanned pregnancy on observable divorce rates, especially for young women.
\end{abstract}

@@

This paper has two main goals. First, I document and quantify the persistent difference in divorce rates of couples who experienced shotgun marriages and how it is pronounced for different population groups. Second, based on these findings I estimate a structural model that matches the observable variation in the data and is capable of identifying importance of different channels that drive this difference.

Combining the empirical variation with the model provides better understanding of people's marriage decisions and general value of marriage, which is being discussed since the the work of \cite{becker1981}. Modern studies emphasize roles of risk-sharing, as in \cite{liseyamada}, general comparative advantage of being a couple as in \cite{chiappori1997} and shared production of public good as in \cite{greenwood}. Having all these mechanisms in place the current work primarily focuses on the last one: children are the crucial part of the value that is created within a couple and therefore are first order issue in how couples are formed. Using the data alone provides pretty limited insights for this problem, and this justifies using a structural model capable of replicating the observed variation. One fundamental question that is possible to address is how large is the value of marriage per se relative to the value of possibility of couple to have children together, and exploiting how people choose to enter inferior quality marriages for the opportunity to raise their child together seems directly relevant.

The idea of exploring outcomes of people who married after they got pregnant is similar in spirit to the literature analyzing teenage pregnancy. From pioneering work of \cite{cardwise} to modern \cite{ashcraft} and \cite{rosenbaum} people were using different data strategies to understand how early birth affect outcomes of women, where more recent works suggest that observable impact is driven majorly be self-selection. My work focuses on a distinct angle of the similar issue: in fact, although large chunk of shotgun marriages happens in early life, they do not appear to be the driver of the results: excluding early marriages or controlling for mother's age at the first birth and at the first marriage does not mitigate differences in marital performance --- shotgun marriages that happen later in life have very similar consequences to the early ones. Finally, my work does not aim to recover the results from the data alone: instead of dealing with the selection empirically I attempt to model it, though focusing mostly on a more narrow issues of marriage performance and couple formation.

Understanding the sources of a variation in marriage performance is not just a household behavior question. For example, divorces have major impact on how people raise their children. The families who experienced shotgun marriages are often overlooked by the current literature focusing on child development. Growing up with a single mother or in a blended family (i.e. with step-parents) is conventionally considered as a predictor of poor early and adult-life outcomes of the children that is reviewed, for instance, by \cite{kearney}. Moreover, work of \cite{chetty} suggests that high shares of single mothers in neighborhoods are very strong predictor of low intergenerational mobility and reproduction of inequality. In these studies those who have two married biological parents are typically not considered to be at risk. However, not all marriages are created equal, and treating all marriages equally may understate inequality: growing up in unstable marriage can be a predictor of ending up with a single mother or in a different household, as well as factor of risk by its own.

@@

%[Paragraph 7: model --- setup and why do we need a model]
These comparisons, however, contain lots of selection issues and unobservable heterogeneity, and to deal with it I set up a structural model that is capable of understanding of anatomy of the result. The model considers marriage as a contract where partners cannot commit not to break up, based on literature on limited commitment such as \cite{kocherlakota}, with dynamic nature as in \cite{marcetmarimon}. The growing branch of literature applies this framework to the collctive household models, so my model is related to such as \cite{voena-1}, \cite{mazzocco} or \cite{shephard}. In the model I endogenize savings, marriage and divorce decisions, fertility decisions and fertility shocks, together with the structure of childcare costs. The last element is crucial in generating heterogeneous response in terms of income or education: as time of wealthier parents is more expensive, unexpected childcare costs create relatively more distress for them. The model is estimated using simulated method of moments to match the key patterns of the data coming from American Community Survey, as well as supplementary sources, including ATUS and CEX for disciplining childcare costs part, and the structure I use fits the data well.

%[Paragraph 8: model --- what generates the result in theory]
The main driving force in the model is an unplanned pregnancy shock. Each period single people meet potential partners with some probability, and if the partner is met, the couple decides whether to marry each other based on a simple bargaining procedure. The potential matches are heterogeneous in partner's characteristics and marriage quality, the latter is modeled as idiosyncratic joint utility gain of couple from being together, that can be interpreted as a ``love shock''.  Before the bargaining decision takes place, partners may or may not get an unplanned pregnancy shock, that proxies for unplanned children appearing in cohabitating unions as I do not model cohabitation explicitly. The shock interfers with bargaining: if the couple made a baby, the woman risks to stay a single mother if the couple does not agree to marry. The conjecture is that in matches that experienced the shock women agree to enter the marriages of lower quality. Importance and even presence of this channel, however, depend on parameters of individual preferences and childcare costs that I estimate structurally. However, as a generic feature, after a pregnancy shock happen, the key reason that women agree to stick to the poor-quality matches is splitting the unexpected costs, and the key reason for men is the access to their children. This suggests that subsidizing childcare or enforcing visitation rights would decrease number of such marriages at the cost of increasing the number of single parents.

%[Paragraph 9: model --- what I get from it, mechanics and anatomy]
The estimation results confirm lower unobserved marriage quality of those who experienced an unplanned pregnancy shock and suggest this to be the major driver of the number that we observe in the data. The model is capable of delivering various decompositions that split observable cross-sectional difference in divorce rates onto effects of age, marriage length, income, wealth and marriage quality. As it turns out, the last factor is the most important contributor to the difference, and its contribution is [of the similar magnitude than the observable difference]. Decomposing the result further, the model predicts that the major part of the difference in divorce rates is driven by couples who would not have married if they did not have a baby (i.e. who complied with the shock), although there is some additional distress of having an unplanned baby on those who would have married anyway.

%[Paragraph 10: model --- counterfactuals and speculative things]
Additionally, the estimated model delivers predictions exploiting interaction between unplanned pregnancies and marriage quality. First, a major part of divorces that happen early in life are seen as driven by unexpected pregnancies, though divorces happening later are not. Second, the general impact of pressure of becoming a single mother on marriage decisions is substantial, and eliminating unexpected pregnancies happening before the marriage decision takes place lowers amount of people married by 25 by [...].
%Third, subsample analysis suggest that the most important driver of stronger impact of shotgun marriages on college graduates relative to non-college graduates can be rationalized by larger weight of child quality in their utility.
%[Last, the model predicts decline in probabilities of unexpected pregnancies since 1980: the model matching fertility / marriage / divorce profiles from Census of 1980 predicts larger chances of unplanned pregnancies among all income and age groups.]

@@

%[Paragraph 12: what I learned about channels]
The main causal insight is that being a single mother is a major threat affecting how people make their marriage decisions, and large and unequally distributed costs of raising children can be a driver in the creation of poor quality marriages. This channel can be of major importance for studying welfare policies subsidizing childcare, young families and single mothers, as well as for enforcing alimony payments, child support and visitation rights in general. This also emphasizes the link between contraception and abortion access and marriages, although this impact is limited as even the wealthiest couples do experience sizable share of shotgun marriages, though this may be cultural rather than economic issue.

@@

\section*{Literature}
%\bibliographystyle{achicago}
\bibliography{bib}
